---
title: "RecFin Data Overview"
author: "Zoë Kitchel"
date: "`r Sys.Date()`"
output: bookdown::html_document2
---

```{r setup,  include = FALSE}
library(data.table)
library(ggplot2)
library(knitr)
library(kableExtra)

```


## How we pulled data

-   Downloaded RecFin data on January 15, 2025 from [recfin
    website](https://reports.psmfc.org/recfin/f?p=601:1:13820808312939:INITIAL:::F_SELECTED_NODE:53&cs=3MMX4GLd8uuMBgMGOzscCAHvH_yUely3YcYGno0ODuz8s3Y2PDfjORospWxOA5ztpEkJ4jVH0U06YeC5-Kvv10g)

Filters applied for download:

-   Agency = California
-   Mode = All (Note, for some reason, there is NO Beach/Bank in
    2023-2024)
-   Water Area = Ocean (\<= 3 miles)
-   Subregion = South California
-   Trip Type = All

Thoughts

-   Not sure why no beach/bank in 2023-2024
    -   Beach bank catch estimates only based on online surveys, maybe
        these haven't been integrated yet?
    -   ALDOS (Angler License Directory Online Survey) is used to
        collect effort data only for beach and bank fishing, night-time
        fishing, and private access fishing that might otherwise go
        unaccounted for.
-   Should we include South district (San Diego, Orange, and LA
    Counties) and Channel district (Ventura and Santa Barbara Counties)?
    or just South?
    -   Ideally, just LA County?
-   Options for units are retained count, retained weight, released
    count, released weight, released dead count, released dead weight,
    total mortality count, total mortality weight
    -   Likely, we're most interested in Total Mortality count and weight (and therefore that's what I include below)

Helpful resources and tips

- [NOAA on CDFW-ALDOS](https://www.fisheries.noaa.gov/s3/2023-09/CDFW-ALDOS-PR1-PR2-Certification-OST-2023-Redacted.pdf)
-   RecFin data is a Catch Estimate
-   1979 - 2003 = Marine Recreational Statistical Survey (MRFSS)
-   2004 - Present = California Recreational Fisheries Survey (CRFS)
-   1936 - Present = Commercial Passenger Fishing Vessel (CPFV)
    operators are required to submit logbooks to CDFW that include
    documentation of each fishing trip
    -   Each logbook entry reports effort (\# anglers and \# hours
        fished) and take (species and number fish caught)
    -   CRFS uses CPFV effort data collected by samplers in the field
        and logbook records to estimate CPFV effort

## Data processing steps
### I load the data from RecFin website, and merge all CSVs into one.

```{r,  echo=FALSE, warning=FALSE}
# List all CSV files in the directory that match CTE001-California pattern
csv_files <- list.files(path = here::here("data", "RecFin"), pattern = "CTE001.*\\.csv$", full.names = TRUE)

# Read and merge all CSV files
recfin_data_merge <- rbindlist(lapply(csv_files, fread), fill = TRUE)

# Write to a new file with updated year
fwrite(recfin_data_merge, file = here::here("data","RecFin","recfin_data_merge.csv"))

```

### I make some small data adjustments to improve usability.

```{r,  echo=FALSE, warning=FALSE}
#Order factor for mode and add return to labels
recfin_data_merge[,mode_label := factor(RECFIN_MODE_NAME, levels = c( "Beach/Bank", "Man-Made/Jetty", "Party/Charter Boats", "Private/Rental Boats"), labels = c( "Beach/\nBank", "Man-Made/\nJetty", "Party/\nCharter Boats", "Private/\nRental Boats"))]
```

### Start by visualizing all RecFin data from 2004-2024


```{r,  echo=FALSE, warning=FALSE, fig.dim = c(7, 8), fig.cap = "RecFin total estimated fishing mortality by count. Note different y-axis scale for each district. Color specifies fishing mode, line type specifies fishing mode location."}
#Add column that differentiates shore and boat
recfin_data_merge[,shore_boat := ifelse(RECFIN_MODE_NAME %in% c("Party/Charter Boats", "Private/Rental Boats"),"Boat-based","Shore-based")]

#Adjust factor order
recfin_data_merge[,shore_boat := factor(shore_boat, levels = c("Shore-based","Boat-based"))]

recfin_data_merge_total_mortality <- recfin_data_merge[,.(SUM_TOTAL_MORTALITY_NUM_ANNUAL = 
                                                             sum(SUM_TOTAL_MORTALITY_NUM, na.rm = T),
                                                            SUM_TOTAL_MORTALITY_MT_ANNUAL = 
                                                             sum(SUM_TOTAL_MORTALITY_MT, na.rm = T)),
                                                       by = .(RECFIN_YEAR, DISTRICT_NAME,RECFIN_MODE_NAME, shore_boat)]

district_labels <- c("Channel District\n(Ventura and Santa Barbara Counties)",
                     "South District\n(San Diego, Orange, and LA Counties)")

#Change district names
recfin_data_merge_total_mortality[,DISTRICT_NAME_LABEL := factor(DISTRICT_NAME, labels = district_labels)]

ggplot(recfin_data_merge_total_mortality) +
    geom_line(aes(x = RECFIN_YEAR, y = SUM_TOTAL_MORTALITY_NUM_ANNUAL/1000, color = RECFIN_MODE_NAME, linetype = shore_boat)) +
    scale_x_continuous(breaks = seq(2005,2025,by = 5), limits = c(2005,2025)) +
    scale_color_manual(values = c("#B276C2","#E28080","#78A7CF","#88C89A")) +
    scale_linetype_manual(values = c("solid","dashed")) +
    facet_wrap(~DISTRICT_NAME_LABEL, ncol = 1, labeller = labeller(district_labels), scales = "free") +
    labs(x = "Year",y = "Total mortality (count in 1000s)", color = "Fishing mode", linetype = "Fishing location") +
      theme_classic()
    
```


```{r,  echo=FALSE, warning=FALSE, fig.dim = c(7, 8), fig.cap = "RecFin total estimated fishing mortality by weight. Note different y-axis scale for each district. Color specifies fishing mode, line type specifies fishing mode location."}
ggplot(recfin_data_merge_total_mortality) +
    geom_line(aes(x = RECFIN_YEAR, y = SUM_TOTAL_MORTALITY_MT_ANNUAL, color = RECFIN_MODE_NAME, linetype = shore_boat)) +
    scale_x_continuous(breaks = seq(2005,2025,by = 5), limits = c(2005,2025)) +
    scale_color_manual(values = c("#B276C2","#E28080","#78A7CF","#88C89A")) +
    scale_linetype_manual(values = c("solid","dashed")) +
    facet_wrap(~DISTRICT_NAME_LABEL, ncol = 1, labeller = labeller(district_labels), scales = "free") +
    labs(x = "Year",y = "Total mortality (metric tonnes)", color = "Fishing mode", linetype = "Fishing location") +
    theme_classic()
```



### Identify most common species



```{r,  echo=FALSE, warning=FALSE}
mortality_totals <- recfin_data_merge[,.(SUM_TOTAL_MORTALITY_NUM_ALLYEARS_1000s = 
                                                             sum(SUM_TOTAL_MORTALITY_NUM/1000, na.rm = T),
                                                            SUM_TOTAL_MORTALITY_MT_ALLYEARS = 
                                                             sum(SUM_TOTAL_MORTALITY_MT, na.rm = T)),
                                                       by = .(SPECIES_NAME)]

#Top species by count
setkey(mortality_totals, SUM_TOTAL_MORTALITY_NUM_ALLYEARS_1000s)
mortality_totals_count <- data.table(tail(mortality_totals,10))

#Top species by weight
setkey(mortality_totals, SUM_TOTAL_MORTALITY_MT_ALLYEARS)
mortality_totals_weight <- data.table(tail(mortality_totals,10))

#Round and format
mortality_totals_weight[,`Total mortality 2004-2024 (metric tonnes)` := round(SUM_TOTAL_MORTALITY_MT_ALLYEARS,0)][,SUM_TOTAL_MORTALITY_NUM_ALLYEARS_1000s := NULL][,SUM_TOTAL_MORTALITY_MT_ALLYEARS := NULL][,rank := c(10:1)]
setnames(mortality_totals_weight, old = "SPECIES_NAME",new = "Species")
setkey(mortality_totals_weight, rank)
mortality_totals_weight[,rank := NULL]

mortality_totals_count[,`Total mortality 2004-2024 (count in 1000s)` := round(SUM_TOTAL_MORTALITY_NUM_ALLYEARS_1000s,0)][,SUM_TOTAL_MORTALITY_NUM_ALLYEARS_1000s := NULL][,SUM_TOTAL_MORTALITY_MT_ALLYEARS := NULL][,rank := c(10:1)]
setnames(mortality_totals_count, old = "SPECIES_NAME",new = "Species")
setkey(mortality_totals_count, rank)
mortality_totals_count[,rank := NULL]

```

```{r,  echo=FALSE, warning=FALSE}
knitr::kable(mortality_totals_count,format = "html", caption = "Top species (by count) across both districts and all fishing modes.") |> row_spec(row = 0, bold = T)
```


```{r,  echo=FALSE, warning=FALSE}
knitr::kable(mortality_totals_weight,format = "html", caption = "Top species (by weight) across both districts and all fishing modes.") |> row_spec(row = 0, bold = T)
```

```{r,  echo=FALSE, warning=FALSE}
#Calculate again, but only for SD, Orange, and LA counties
SouthDistrict_estimated_mortality <- recfin_data_merge[DISTRICT_NAME == "South (San Diego, Orange And Los Angeles Counties)"]

SouthDistrict_mortality_totals <- SouthDistrict_estimated_mortality[,.(SUM_TOTAL_MORTALITY_NUM_ALLYEARS_1000s = 
                                                             sum(SUM_TOTAL_MORTALITY_NUM/1000, na.rm = T),
                                                            SUM_TOTAL_MORTALITY_MT_ALLYEARS = 
                                                             sum(SUM_TOTAL_MORTALITY_MT, na.rm = T)),
                                                       by = .(SPECIES_NAME)]

#Top species by count
setkey(SouthDistrict_mortality_totals, SUM_TOTAL_MORTALITY_NUM_ALLYEARS_1000s)
SouthDistrict_mortality_totals_count <- data.table(tail(SouthDistrict_mortality_totals,10))

#Top species by weight
setkey(SouthDistrict_mortality_totals, SUM_TOTAL_MORTALITY_MT_ALLYEARS)
SouthDistrict_mortality_totals_weight <- data.table(tail(SouthDistrict_mortality_totals,10))

#Round and format
SouthDistrict_mortality_totals_weight[,`Total mortality South District 2004-2024 (metric tonnes)` := round(SUM_TOTAL_MORTALITY_MT_ALLYEARS,0)][,SUM_TOTAL_MORTALITY_NUM_ALLYEARS_1000s := NULL][,SUM_TOTAL_MORTALITY_MT_ALLYEARS := NULL][,rank := c(10:1)]
setnames(SouthDistrict_mortality_totals_weight, old = "SPECIES_NAME",new = "Species")
setkey(SouthDistrict_mortality_totals_weight, rank)
SouthDistrict_mortality_totals_weight[,rank := NULL]

SouthDistrict_mortality_totals_count[,`Total mortality South District 2004-2024 (count in 1000s)` := round(SUM_TOTAL_MORTALITY_NUM_ALLYEARS_1000s,0)][,SUM_TOTAL_MORTALITY_NUM_ALLYEARS_1000s := NULL][,SUM_TOTAL_MORTALITY_MT_ALLYEARS := NULL][,rank := c(10:1)]
setnames(SouthDistrict_mortality_totals_count, old = "SPECIES_NAME",new = "Species")
setkey(SouthDistrict_mortality_totals_count, rank)
SouthDistrict_mortality_totals_count[,rank := NULL]

```

```{r,  echo=FALSE, warning=FALSE}
knitr::kable(SouthDistrict_mortality_totals_count, format = "html", caption = "Top species (by count) for all fishing modes in South District (LA, SD, OC Counties).") |> row_spec(row = 0, bold = T)
```

```{r,  echo=FALSE, warning=FALSE}
knitr::kable(SouthDistrict_mortality_totals_weight, format = "html", caption = "Top species (by weight) for all fishing modes in South District (LA, SD, OC Counties).") |> row_spec(row = 0, bold = T)
```

```{r,  echo=FALSE, warning=FALSE}
#Calculate again, but only for SD, Orange, and LA counties
SouthDistrict_ShoreMode_estimated_mortality <- recfin_data_merge[DISTRICT_NAME == "South (San Diego, Orange And Los Angeles Counties)" & shore_boat == "Shore-based"]

SouthDistrict_ShoreMode_mortality_totals <- SouthDistrict_ShoreMode_estimated_mortality[,.(SUM_TOTAL_MORTALITY_NUM_ALLYEARS_1000s = 
                                                             sum(SUM_TOTAL_MORTALITY_NUM/1000, na.rm = T),
                                                            SUM_TOTAL_MORTALITY_MT_ALLYEARS = 
                                                             sum(SUM_TOTAL_MORTALITY_MT, na.rm = T)),
                                                       by = .(SPECIES_NAME)]

#Top species by count
setkey(SouthDistrict_ShoreMode_mortality_totals, SUM_TOTAL_MORTALITY_NUM_ALLYEARS_1000s)
SouthDistrict_ShoreMode_mortality_totals_count <- data.table(tail(SouthDistrict_ShoreMode_mortality_totals,10))

#Top species by weight
setkey(SouthDistrict_ShoreMode_mortality_totals, SUM_TOTAL_MORTALITY_MT_ALLYEARS)
SouthDistrict_ShoreMode_mortality_totals_weight <- data.table(tail(SouthDistrict_ShoreMode_mortality_totals,10))

#Round and format
SouthDistrict_ShoreMode_mortality_totals_weight[,`Total shore-based mortality South District 2004-2024 (metric tonnes)` := round(SUM_TOTAL_MORTALITY_MT_ALLYEARS,0)][,SUM_TOTAL_MORTALITY_NUM_ALLYEARS_1000s := NULL][,SUM_TOTAL_MORTALITY_MT_ALLYEARS := NULL][,rank := c(10:1)]
setnames(SouthDistrict_ShoreMode_mortality_totals_weight, old = "SPECIES_NAME",new = "Species")
setkey(SouthDistrict_ShoreMode_mortality_totals_weight, rank)
SouthDistrict_ShoreMode_mortality_totals_weight[,rank := NULL]

SouthDistrict_ShoreMode_mortality_totals_count[,`Total shore-based mortality South District 2004-2024 (count in 1000s)` := round(SUM_TOTAL_MORTALITY_NUM_ALLYEARS_1000s,0)][,SUM_TOTAL_MORTALITY_NUM_ALLYEARS_1000s := NULL][,SUM_TOTAL_MORTALITY_MT_ALLYEARS := NULL][,rank := c(10:1)]
setnames(SouthDistrict_ShoreMode_mortality_totals_count, old = "SPECIES_NAME",new = "Species")
setkey(SouthDistrict_ShoreMode_mortality_totals_count, rank)
SouthDistrict_ShoreMode_mortality_totals_count[,rank := NULL]

```

```{r,  echo=FALSE, warning=FALSE}
knitr::kable(SouthDistrict_ShoreMode_mortality_totals_count, format = "html", caption = "Top species (by count) for shore-based fishing modes in South District (LA, SD, OC Counties).") |> row_spec(row = 0, bold = T)

```

```{r,  echo=FALSE, warning=FALSE}
knitr::kable(SouthDistrict_ShoreMode_mortality_totals_weight, format = "html", caption = "Top species (by weight) for shore-based fishing modes in South District (LA, SD, OC Counties).") |> row_spec(row = 0, bold = T)
```



### Visualize time series of most common species in South District caught using shore-based methods (Beach/Pier/Jetty).

```{r,  echo=FALSE, warning=FALSE}
SouthDistrict_ShoreMode_TopSpp <- unique(c(SouthDistrict_ShoreMode_mortality_totals_weight$Species, SouthDistrict_ShoreMode_mortality_totals_count$Species))

```


```{r,  echo=FALSE, warning=FALSE}
#Sum if not in top for shore-based fishing in South District, plot time-series of counts and weights as areas.
SouthDistrict_ShoreMode_estimated_mortality[,SPECIES_NAME_TOP_SOUTHDISTRICT_SHOREBASED := ifelse(SPECIES_NAME %in% SouthDistrict_ShoreMode_TopSpp,SPECIES_NAME,"Other")]

#Sum across new spp designations
#Unique rows for total count and weight of each species per year per mode
SouthDistrict_ShoreMode_estimated_mortality.r <- SouthDistrict_ShoreMode_estimated_mortality[,.(SUM_TOTAL_MORTALITY_NUM_ANNUAL = 
                                                             sum(SUM_TOTAL_MORTALITY_NUM, na.rm = T),
                                                            SUM_TOTAL_MORTALITY_MT_ANNUAL = 
                                                             sum(SUM_TOTAL_MORTALITY_MT, na.rm = T)),
                                                       by = .(RECFIN_YEAR,SPECIES_NAME_TOP_SOUTHDISTRICT_SHOREBASED, shore_boat, mode_label, RECFIN_MODE_NAME)]


#Set order of spp for plotting
most_common_spp_wother <- c(rev(SouthDistrict_ShoreMode_TopSpp),"Other")

SouthDistrict_ShoreMode_estimated_mortality.r[,SPECIES_NAME_TOP := factor(SPECIES_NAME_TOP_SOUTHDISTRICT_SHOREBASED,levels = most_common_spp_wother)]

#Color palette
pal <- c("#FF6B6B",
      "#6B8E23",
      "#4682B4",
      "#DDA0DD",
      "#FFA07A",
      "#20B2AA",
      "#FF8C00",
      "#9370DB",
      "#32CD32",
      "#DC143C",
      "#8B4513", 
      "#00CED1",
      "#FFD700",
      "#FF1493",
      "black",
      "darkgrey")


#Count plot
SouthDistrict_ShoreMode_TimeSeries_AreaPlot_Count <- ggplot(data = SouthDistrict_ShoreMode_estimated_mortality.r, aes(x = RECFIN_YEAR, SUM_TOTAL_MORTALITY_NUM_ANNUAL/1000,
                                                 fill = SPECIES_NAME_TOP, label = SPECIES_NAME_TOP, color = SPECIES_NAME_TOP)) +
  geom_area() +
  scale_fill_manual(values=pal) +
  scale_color_manual(values=pal) +
    scale_x_continuous(breaks = seq(2005,2025,by = 5), limits = c(2005,2025)) +
  scale_y_continuous(expand = c(0,0)) +
    facet_wrap(~RECFIN_MODE_NAME, scales = "free", ncol = 1)+
        labs(x = "Year",y = "Total mortality (count in 1000s)", color = "Species", fill = "Species",label = "Species") +
    theme_classic() +
    theme(legend.position = "top", legend.direction = "horizontal")

#Weight plot
SouthDistrict_ShoreMode_TimeSeries_AreaPlot_Weight <- ggplot(data = SouthDistrict_ShoreMode_estimated_mortality.r, aes(x = RECFIN_YEAR, SUM_TOTAL_MORTALITY_MT_ANNUAL,
                                                 fill = SPECIES_NAME_TOP, label = SPECIES_NAME_TOP, color = SPECIES_NAME_TOP)) +
  geom_area() +
  scale_fill_manual(values=pal) +
  scale_color_manual(values=pal) +
    scale_x_continuous(breaks = seq(2005,2025,by = 5), limits = c(2005,2025)) +
  scale_y_continuous(expand = c(0,0)) +
    facet_wrap(~RECFIN_MODE_NAME, scales = "free", ncol = 1)+
        labs(x = "Year",y = "Total mortality (metric tonnes)", color = "Species", fill = "Species",label = "Species") +
    theme_classic() +
    theme(legend.position = "top", legend.direction = "horizontal")

```


```{r, echo=FALSE, warning=FALSE, fig.dim = c(7, 8), fig.cap = "RecFin total estimated mortality count for shore-based fishing in the South District. All species not in the top ten by either count or weight are grouped as Other. Note different y-axis scale for mode. Color identifies species."}
SouthDistrict_ShoreMode_TimeSeries_AreaPlot_Count
```



```{r, echo=FALSE, warning=FALSE, fig.dim = c(7, 8), fig.cap = "RecFin total estimated mortality weight for shore-based fishing in the South District. All species not in the top ten by either count or weight are grouped as Other. Note different y-axis scale for mode. Color identifies species."}
SouthDistrict_ShoreMode_TimeSeries_AreaPlot_Weight
```


```{r,  echo=FALSE, warning=FALSE}

#Species name label edit (add returns)
SouthDistrict_ShoreMode_estimated_mortality.r[,SPECIES_NAME_TOP_longlabel := gsub(" ","\n",SPECIES_NAME_TOP)]

#Count plot
SouthDistrict_ShoreMode_TimeSeries_LinePlot_Count <- ggplot(data = SouthDistrict_ShoreMode_estimated_mortality.r[SPECIES_NAME_TOP != "Other"], aes(x = RECFIN_YEAR, SUM_TOTAL_MORTALITY_NUM_ANNUAL/1000, label = SPECIES_NAME_TOP, color = SPECIES_NAME_TOP)) +
  geom_line(linewidth = 1) +
    geom_rug(sides = "b") +
  scale_fill_manual(values=pal) +
  scale_color_manual(values=pal) +
    scale_x_continuous(breaks = seq(2005,2025,by = 5), limits = c(2005,2025)) +
  scale_y_continuous(expand = c(0,0)) +
    facet_grid(SPECIES_NAME_TOP_longlabel~RECFIN_MODE_NAME, scales = "free")+
labs(x = "Year",y = "Total mortality (count in 1000s)", color = "Species", fill = "Species",label = "Species") +
    theme_classic() +
        theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.05),  # Adds axis lines
    strip.text.y = element_text(size = 6)  # Makes y-axis strip text smaller
  )

#Weight plot
SouthDistrict_ShoreMode_TimeSeries_LinePlot_Weight <- ggplot(data = SouthDistrict_ShoreMode_estimated_mortality.r[SPECIES_NAME_TOP != "Other"], aes(x = RECFIN_YEAR, SUM_TOTAL_MORTALITY_MT_ANNUAL, label = SPECIES_NAME_TOP, color = SPECIES_NAME_TOP)) +
  geom_line(linewidth = 1) +
    geom_rug(sides = "b") +
  scale_fill_manual(values=pal) +
  scale_color_manual(values=pal) +
    scale_x_continuous(breaks = seq(2005,2025,by = 5), limits = c(2005,2025)) +
    facet_grid(SPECIES_NAME_TOP_longlabel~RECFIN_MODE_NAME, scales = "free")+
        labs(x = "Year",y = "Total mortality (metric tonnes)", color = "Species", fill = "Species",label = "Species") +
    theme_classic() +
    theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.05),  # Adds axis lines
    strip.text.y = element_text(size = 6)  # Makes y-axis strip text smaller
  )

```


```{r, echo=FALSE, warning=FALSE,message = FALSE, fig.dim = c(7, 8), fig.cap = "Total estimated mortality (count) for most commonly harvested species (by either weight or count) using shore-based fishing methods in the South District. Data from RecFin. Note different y-axis scale across species. Color identifies species."}
SouthDistrict_ShoreMode_TimeSeries_LinePlot_Count

```

```{r, echo=FALSE, warning=FALSE, fig.dim = c(7, 8),message = FALSE, fig.cap = "Total estimated mortality (weight) for most commonly harvested species (by either weight or count) using shore-based fishing methods in the South District. Data from RecFin. Note different y-axis scale across species. Color identifies species."}

SouthDistrict_ShoreMode_TimeSeries_LinePlot_Weight
```


I'm not immediately worried about changes in ID over time for any top species. It would be different if we were looking at rockfish, but no rockfish caught from shore.